Predicting the PCG PSeudo-Random Number Generator In Practice

### 2020-05-08 article/main.tex
up-to-date
ajout du rapport de projet sans unknown increment (qui est juste un algo faux)

@@

\section{Introduction}

@@

\section{Warm-Up: Prediction with Known Increment }

>>>>

%Ceux qu'on utilisait dans le projet :

\usepackage{algorithm}
\usepackage{algorithmic}

@@

\section{Introduction} %ce qu'on avait écrit pour le projet, peut-être pas incroyable...

Today's world has become a data driven world. Having access unpredictable data is invaluable when considering security problem. For example, security keys have to be generated randomly whether we are talking about RSA keys for secure communication or a cipher key for encrypting data. For this purpose, random data can be generated by pseudo-random generators, that rely on a predetermined mathematical structure to output seemingly random data. However, by knowing the internal structure and state of the generator, it is possible to reconstruct the entire string of data.\\

The goal of our project is to study the flaws and try to cryptanalyse pseudo-random generators from the Permuted Congruential Generator (PCG) family, designed by Melissa E. O'Neill.\cite{melissapaper} Indeed, predicting a sequence from these generators is said to be "challenging"\cite{melissaweb} by their author, but this statement is not backed by any studies. Furthermore, PCG generators are very close to another well known family of generators, the Truncated Linear Congruential Generators, for which a certain amount of viable prediction algorithms are known. The combination between being highly commended (it achieved 5th place out of 29 in a recent performance test survey\cite{survey}) and a well presented website\cite{melissaweb} giving an feeling of professionalism can lead to the impression that PCG is a sufficiently secure family of generators. Given the lack of studies on the subject, there is a problem if it is used in critical pieces of software. This is why we will attempt to find flaws in PCG random generators.\\

Some attacks against PCG have been carried out by Sebastiano Vigna\cite{vignaweb}, but this didn't yield any significant results, given that he only tried to break an incomplete version in a fairly straightforward way. He worked on a very specific version of PCG. He considers the result without modulo, which is a big simplification. Moreover, he didn't study the 128 bits long generator, but only the 64 bits long\cite{vignacode}.\\

We focused on the general version of PCG, which happens to be very similar to the generator implemented by the function \texttt{random} of the C library. Both are based on a well known family of generators, the Linear Congruential Generators (LCG), but with the PCG one also using a non linear function and a rotation on the output.\\

We are looking for a possible way to reduce this generator to an LCG, or a truncated LCG, because there are known algorithms to predict them. Because we aren't studying the same exact version as Sebastiano Vigna, his findings aren't very relevant to us.\\

We decided to study the default version of the largest available generator of the 0.94 version of the PCG family, which has an internal state of 128 bits, named PCG128. We will mathematically analyse the algorithm to uncover weaknesses, and if methods for the LCG family can be translated to the PCG family.

@@

\subsection{LCG algorithm}

The PCG algorithm we studied is based on the Linear Congruential Generators algorithm. The LCG algorithm is simply based on a congruential multiplication. This type of generator is based on four values: an internal state $S_{i}$, a multiplier $A$, an increment $C$ and the size $M$. Generally, to initialize a generator of a given size, a seed value is provided as the first internal state, and a default multiplier and increment are used. We will study the case where $M$ and $A$ are known and where $M$ is a power of 2.\\

Calling the generator results in calculating the next internal state $S_{i+1} = A \times S_i + C \mod{M}$ and outputting $S_{i+1}$ as data.

This type of generator is trivially predictable because, the internal state is given to us at each output, and in the event that $C$ is unknown, we can still deduce it with two consecutive outputs: $C = S_{i+1} - A \times S_i$.\\

A better variant of the standard LCG is the Truncated LCG, where the output consists only of the higher bits of the internal state $S_i$. This way the entire internal state isn't accessible. Returning the lower bits is not interesting because as we can understand easily and will prove later, lower bits of a state only depend on the lower bits of the previous state.

\subsection{PCG algorithm}

The PCG algorithm we studied is based on the LCG algorithm. To start, here are a few definitions:
\begin{itemize}
    \item let $k$ be the size of our generator, in this case $k = 128$ bits;
    \item let $M$ be the modulus with $M = 2^k$;
    \item let $S_i$ be the internal state at the $i$-th iteration and $S$ be the list of $S_i$;
    \item let $X_i$ be the output at the $i$-th iteration of the generator and $X$ be the list of $X_i$;
    \item let $lowS_i$ be the $k/2$ least significant bits of $S_i$;
    \item let $upS_i$ be the $k/2$ most significant bits of $S_i$;
    \item let $rot_i$ be the 6 most significant bits of $S_i$;
    \item let $nbiter$ be the number of consecutive outputs for prediction purposes.
\end{itemize}

The internal state can be redefined as:
\begin{equation}
    S_i = 2^{k/2} \times upS_i + lowS_i 
\end{equation}

The 6 most significant bits, $rot_i$, of the internal state are used as a value to determine the rotation in the PCG algorithm. Let \texttt{rotate} be the function that performs said rotation (Figure \ref{pcg128out}).\\
The rotation is a simple exchange of bits, where a $k/2$ bit value is split into 2 parts: one of $rot_i$ bits and the other in $(k/2) - rot_i$ bits. The two parts are then swapped, and the new value is returned.

This generator is initialised with a given $k$ bit internal state $S_0$ (of size 128 bits in the case of PCG128), and optionally with an $k$ bit increment $C$. If no increment is given, the algorithm will use a known default increment. For each output of the generator, the algorithm will perform the following calculations" (Figure \ref{pcg128out}):\\
\begin{align}
    X_{i} &= \mathtt{rotate}(upS_i\ \mathtt{XOR}\ lowS_i)\\
    S_{i+1} &= A \times S_{i} + C \mod{2^k}
\end{align}
In the case of PCG128, $A$ is a default increment of size 125 bits prime with $2^k$.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.75\linewidth]{pictures/PCG128.png}
    \caption{PCG128 : Output process}
    \label{pcg128out}
\end{figure}

Given the nature of the generator, the following values are known at every step:
\begin{itemize}
    \item the size of the generator $k$;
    \item the list of outputs $X$;
    \item the multiplier $A$;
    \item and eventually $C$, if the default increment was used. 
\end{itemize}

To break this generator, we had to differentiate between cases when the default $C$ was used or not. It was also interesting to note what was the minimum number of $X_i$ needed to reliably attack it.

\subsection{Brute-Force Attack with known $C$}
In the case where $C$ is known, if we find $S_0$, we can calculate every $S_i$ with the given formula, and predict all the following outputs of the generator. The naive way is by doing a comprehensive research of $upS_0$, and then calculating $lowS_0$ with the formula :\\
\begin{equation}
    lowS_0 = upS_0\ \mathtt{XOR}\ \mathtt{unrotate}(X_0)
\end{equation}
\texttt with {unrotate} being the inverse function of \texttt{rotate}.\\

If our guessing of $upS_0$ is correct, we can obtain the correct $X_1$ with $S_1$ deduced from the initial $S_0$. The preceding is a naive method for predicting PCG generator in $2^{64}$ operations. But we can have other values that verify that test. This will not return $S_0$ but a very short list of possibles $S_0$, with the actual one in it. We just need to test the following outputs to find the real $S_0$. This is yet a very slow way to guess the internal state, so we need a better solution.

\subsection{Brute-Force Attack with unknown $C$}

In case $C$ is unknown, retrieving $S_0$ is harder. The naive algorithm gives us an answer within $2^{128+6}$ simple operations whit the following algorithm:
\\
\begin{algorithmic}[!h]
\FOR{all possibilities of $lowC$ and $upS_0$}
    \STATE \texttt{unrotate}($X_0$)
    \STATE $lowS_0 = upS_0\ \mathtt{XOR}\ X_0$
    \STATE $lowS_1 = lowC + lowS_0 \times A \mod{2^{k/2}}$
    \FOR{all possibilities of $rot_1$}
        \STATE $\mathtt{unrotate}(X_1)$
        \STATE $upS_1 = lowS_1\ \mathtt{XOR}\ X_1$
        \STATE $C = S_1 - A \times S_0 \mod{2^{k}}$
        \STATE $S_2 = S_1 \times A + C$
        \IF {$S_2$ is coherent with  $X_2$}
            \STATE print($C$,$S_0$)
        \ENDIF
    \ENDFOR
\ENDFOR
\end{algorithmic}

This algorithm reveals us $C$ and a list of outputs with $S_0$ in it. The entry is 256 bits long, and we used 3 64 bits long outputs of the generator. There will be about $2^{64}$ possible $S_0$. This is why we need a couple other outputs to get the real $S_0$. Sorting out which is the correct $S_0$ can be done afterwards, and won't affect the complexity. But given the complexity of this algorithm, it is even more unrealistic to solve problem in this fashion then the previous one where $C$ was known.

\subsection{Reduction to a truncated LCG model}
By exploiting the nature of this generator, it becomes apparent that some values can be easily guessed. We already saw that if we guess $rot_i$, then retrieving $X_i$ becomes a simple \texttt{XOR} operation, and in turn, if we have some knowledge of a particular bit of $lowS_i$ (respectively $upS_i$), we can deduce the equivalent bit from $upS_i$ (respectively $lowS_i$).\\

Moreover, there is an interesting proprity concerning the least significant bits of the internal state. Indeed, if we define $lowC$ as the 64 least significant bits of $C$ ($lowC = C \mod{2^{k/2}})$, we have:

\begin{align}
    S_{i+1} &= A \times S_i + C \mod{2^k}\\
    upS_{i+1} \times 2^{k/2} + lowS_{i+1} &= A \times upS_i \times 2^{k/2} + A \times lowS_i + C \mod{2^k} \\
    lowS_{i+1} &= A \times lowS_i + lowC \mod 2^{k/2}
\end{align}
We can see that $lowS_{i+1}$ is independent from $upS_i$. That mean that knowing $lowS_i$, we can deduce all $lowS_j$ with $j>i$. This relationship is used in the Brute-Force attack with known $C$, in the previous section.
We can see on (Figure \ref{pcg128goodbits})%figure
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.70\linewidth]{pictures/deducingPCG128.png}
    \caption{PCG128 : Relevant bits of the outputs}
    \label{pcg128goodbits}
\end{figure}
that we can extract a lot of information by brute forcing certain well chosen bits. We defined $known\_up$ (respectively $known\_low$) the number of upper (respectively lower) bits searched by brute force.
\newpage

\section{Prediction of truncated LCG algorithms}

\subsection{Different solutions for Truncated Linear Congruential Algorithms}

The truncated LCG are not used anymore because there exists well known solutions for predicting their data stream in polynomial time.\\

The first article we found was written by Joan Boyar\cite{Boyar1989}. The subject of her study was a special case of truncated LCG, with $A$, $C$ and $M$ unknown and has access to a very large part of the internal state. This made it so that we couldn't use her findings, because guessing a large part of the internal state wouldn't be cost-effective. However, her study included a very broad state of the art, explaining the specifics of many algorithms, as well as describing in which circumstances they should be used.\\

The algorithm of Frieze \textit{et al}\cite{Frieze} solves the case where $M$ and $A$ are known, $M$ being a power of two and, possibly with a significant portion of the internal state unknown. The known part is the upper bits of the state. This algorithm is coherent with our problem, this being the reason why we focused on this one in particular. However, this algorithm won't work for some particular multipliers. But this problem doesn't concern us, because the default multiplier of \texttt{PCG128} is not one of them.\\

\subsection{Frieze \textit{et al} algorithm}

The algorithm we studied uses lattices, and more specifically use the\break Lenstra–Lenstra–Lovász lattice basis reduction algorithm (LLL algorithm)\cite{LLLpastack}. A large part of this algorithm is very well explained on Stack Overflow\cite{LLLstack}.
We will detail the entire algorithm here.\\

A lattice in $\mathbb{R}^n$ is a subgroup of the additive group $(\mathbb{R}^n,+)$, isomorphic to $(\mathbb{Z}^n,+)$. The LLL algorithm is an approximation algorithm of the reduction of a lattice basis. It returns a reduced basis, with shorter vectors than the ones given. These vectors are a good approximation of the basis with the smallest determinant.\\

To use this algorithm, we reduce the sequence of $S_i$ to a geometric sequence $S'$ and, given the relationship between $S$ and $S'$, we can retrieve $S$. There are different methods to achieve this result and we selected two of them in our optimised algorithms. Consider the following geometric sequence:
\begin{equation}
    S'_i=\alpha ^i \times S'_0 \mod{M}
\end{equation}
With $S'$ as the vector made of all the $S'_i$ of length $nbiter$, and $A$ blocked, the $S'$ vectors are the vectors of the lattice generated by the matrix :

\begin{equation}
L=
\begin{pmatrix} 
M & 0 & & \dots & 0\\
\alpha & -1 & 0 & \dots & 0\\
\alpha ^2 & 0 & -1 & \dots & 0\\
\vdots & & & \ddots & \vdots\\
\alpha ^{nbiter - 1} & 0 & 0 & \dots & -1
\end{pmatrix}
\end{equation}

%Beau "ell" : $\ell$

Let $Y$ be the most significant bits of the internal states of LCG, and $Z$ be the lower bits, with $S = Y + Z$. We already know $Y$ from the outputs and we are searching for $Z$. 
\begin{itemize}
    \item Let $Y'$ be the upper bits of $S'$;
    \item Let $Z'$ be the lower bits;
    \item Let $L'$  be the reduced matrix returned by the LLL algorithm for $L$ ($L' = LLL(L)$);
\end{itemize}

By definition of $L$, we have :
\begin{align}
    L \times S' &=0 \mod{M}\\
    L' \times S' &= 0 \mod{M}\\
    L' \times (Y' + Z') &= M \times \ell\ \text{($\ell$ unknown integer vector)}\\
    L' \times Z' &= M \times \ell - L' \times Y'
\end{align}
We have $M = 2^k$. L' coefficients are small. We want to find the actual value of $\ell$.\\

We suppose we have every coordinate of $L' \times Z'$ smaller than $M/2$. Then every coordinate of $M \times \ell - L' \times Y'$ is smaller than $M/2$ and $M \times \ell$ is the closest vector of $M$ multiples from $L' \times Z'$. We have $\ell$ is the closest integer vector from $(B \times y)/M$.\\

The only unknown variable left is $Z'$ and :
\begin{equation}
    Z' = L'^{-1} \times (M \times k) - Y' 
\end{equation}
Now we know $Z'$, and the integer $S' \mod{M}$, and finally, $S \mod{M}$ the internal state we were looking for.\\

\subsection{Application to our generator}
The $L$ matrix for our problem is :
\begin{equation}
    L=
\begin{pmatrix}
1208925819614629174706176 & 0 &  0 &  0 &  0 &  0 &  0 &  0 &  0 & 0\\
442210214475561085761093 & -1 &  0 &  0 &  0 & 0 &  0 &  0 &  0 &  0\\
548907270092662740397721 & 0 & -1 &  0 &  0 & 0 &  0 &  0 &  0 &  0\\
68528166612430993102141 & 0 &  0 & -1 &  0 & 0 &  0 &  0 &  0 &  0\\
579003058182980434679665 & 0 &  0 &  0 & -1 & 0 &  0 &  0 &  0 &  0\\
907267876411637163239285 & 0 &  0 &  0 &  0 & -1 &  0 &  0 &  0 &  0\\
971193525585951128490121 & 0 &  0 &  0 &  0 & 0 & -1 &  0 &  0 &  0\\
423043601822819931331309 & 0 &  0 &  0 &  0 & 0 &  0 & -1 &  0 &  0\\
722352710310758831226849 & 0 &  0 &  0 &  0 & 0 &  0 &  0 & -1 &  0\\
239218985093571541740965 & 0 &  0 &  0 &  0 & 0 &  0 &  0 &  0 & -1
\end{pmatrix}
\end{equation}
Each coefficient is about 128 bit long. Here is the $L'$ (using the LLL algorithm of Sagemath \cite{sagemath}):
\begin{equation}
\begin{pmatrix}
    L' =
    77 & -105 & -115 &  17 & -108 & -77  & -7 &  50 & 107 &  17\\
     50  & 80 & 9 & -90 &  138 & -104  & 26  & 90 & -17 & -134\\
    14 & -42 & 130 & -67 & -31 & -172 & 122 & -76 & -32 &  14\\
    -61 &  97 &  70 & 161  & -4 & -229 & -23 & -44  & -1 & -26\\
    118  & 33 & -65 &  -3 &  83 & -176 & -187 & -70 &  25 &  34\\
    111 & -151 & -73 & 115 &  36 & 200 &  79  & -1 & -14 & -82\\
    94  & 15 & -18 & -36 & 142 & -47 &  32 & -188 & 161 & -127\\
    126 & -10 & 129 & 173 & -71  & 61 & -137 &  69 & 7 &  77\\
    174 & -62 &  58  & 39 &  79 & -133 &  93 & 115 &  24 & 181\\
    48 & 240 &  -2 & 114 & -15 &  37 & 145 & -100 &  24 & -19\\
\end{pmatrix}
\end{equation}
This algorithm is clearly really efficient, the coefficients of $L'$ are really small. Its bigger coefficient is $240$. Its smaller than $2^8 = 256$. In our case $M = 2^{64 + known\_low}$. In our tests, we had $known\_low = 16$. We need $Z'$ to be less than 72 bits long, which isn't hard to attain.

\section{Warm-Up: Prediction with Known Increment }
\subsection{Reduction to a geometric sequence and validation of our values}

$C$ is known. We are looking for a geometric sequence $S'$ with which we can validate our brute-forced bits and extract $S_0$. Let $polC$ be the vector where $polC_i = \sum_{j=0}^{i} C \times A^j$ and $S'$ where $S'_i = S_i - polC_i$. We have :
\begin{align}
    S'_{i+1} &= S_{i+1} - polC_{i+1}\\
    S'_{i+1} &= A \times S_i + C - polC_{i+1}\\
    S'_{i+1} &= A \times S_i + C - (C + A \times C + ...+ A^{i} \times C)\\
    S'_{i+1} &= A \times (S'_i + polC_i) - A \times (C + ... + A^{i-1} \times C)\\
    S'_{i+1} &= A \times S'_i + A \times polC_i - A \times polC_i\\
    S'_{i+1} &= A \times S'_i
\end{align}

We have a convenient $S'$ that can be used in the algorithm presented in the previous section and, there exist a simple bijection between $S'$ and $S$. However, its important to notice that $S'$ will not be computed modulo $M = 2^k$ but modulo $N = 2^{k/2 + known\_low}$, because we are doing an comprehensive research of the middle bits of $S$ and not the upper ones (except for the 6 bits of $rot$). This is not a problem, because when we have the values of $X$, and $lowS$, we can always obtain $upS$ and ultimately $S$.\\

Using this bijection, we can validate if our chosen values are correct. Indeed, after finding $Z'$ through the LLL algorithm, we can compute $S'_0$ and $S'_1$ modulo $2^{k/2 + known\_low}$, then $S_0$ and $S_1$ modulo $2^{k/2 + known\_low}$, and finally $S_1$ and $S_0$. We verify if $S_1$ is coherent with $S_0$. If it is, then we have guessed the correct values for our generator, if not, we then consider this another value as our next test.

\newpage

\subsection{Pseudo-Algorithm}

\begin{algorithm}[h!]
\caption{Algorithm with known $C$}
\begin{algorithmic}
\REQUIRE $X$ the list of outputs
\STATE computation of the $LLL$ matrix
\STATE computation of $polC$
\FOR{ all $rot_i$ for $i$ in range($nbiter$) }
    \FOR{i in range($nbiter$)}
        \STATE \texttt{unrotate}($X_i$)
    \ENDFOR
    \FOR{all $W_0$}
        \STATE calculate the other $W_i$ with $W_0$ and $C$
        \STATE calculate $Y$ using $W_i\ \mathtt{XOR}\ X_i$ and $rot_i\ \mathtt{XOR}\ X_i$
        \STATE $W' = W - polC$ (we subtract only the corresponding bits of $C$)
        \STATE $Y' = Y - polC$
        \STATE calculation of $Z'$ due to LLL algorithm
        \IF{$2^{k/2-known\_up}>Y'>=2^{known\_low}$}
            \IF{$Z$ is correct}
                \STATE return the corresponding $S_0$
            \ENDIF
        \ENDIF
    \ENDFOR
\ENDFOR
\end{algorithmic}
\end{algorithm}

\subsection{Performance test results}
Our code is implemented in Sagemath. This choice was made because of the development advantages it offered: to be able to work with arbitrarily large integers and have access to a built-in LLL function.\\

The number of bits obtained by exhaustive research, as we have seen in the first section, is $nbiter \times known\_up + known\_low$. The most efficient way should be to only guess the lower bits ($known\_up = 0$). But we need the 6 bit value of $rot_i$ in our algorithm. So we kept $known\_up = 6$ and searched for the best equilibrium between $nbiter$ and $known\_low$. \\

Moreover, we also need this algorithm to be reliable. That is a concern because we know that there is a rounding error caused by an Euclidian Division. Depending on the error, it is possible that the algorithm doesn't find the correct $Z$ even if $Y$ and $W$ are correct (with $W$ the lower bits of $S$ and $Y$ the middle bits of $S$ found by \texttt{XOR}). \\

We tested the reliability and the efficiency of the algorithm. By trial and error, we ended up finding two interesting settings :
\begin{itemize}
    \item $known\_low = 16$ and $nbiter = 4$ : we have a $2^{40}$ brute-force algorithm. When supplying it with PCG128 outputs, the rate at which the algorithm detects it and obtains the correct $S_0$ is of 90\%.
    \item $known\_low = 15$ and $nbiter = 4$ : we have a $2^{39}$ brute-force algorithm. When supplying it with PCG128 outputs, the rate at which the algorithm detects it and obtains the correct $S_0$ is of 70\%.
\end{itemize}

In any case, if we supply a random output, the algorithm will detect this and will never return a false $S_0$. When not finding $S_0$, the default answer is to disqualify the input as a PCG data steam.\\

Unfortunately, we didn't get to run our entire solution in a real time environment, because with our personal equipment, and not fully optimized code, we estimated that it would have take more or less ten years to obtain an answer, which is obviously too long.
The local environment at the university didn't provide the right tools (SageMath), which complexified the development process of sharing the computation load between multiple computers. However, there are some methods and coding improvements to reduce the time needed, as we will discuss in the next section.
